### 2016-09-14 Chapters/MathematicalResults.tex
Save the day's work

This appendix collects some results in the field of functional analysis and
boundary integral equations. No proofs or examples are given, the interested
reader is referred to the monographs by \citeauthor{Ern2004-oo},
\citeauthor{Hsiao2008-xb} and \citeauthor{Sauter2011-an}\autocite{Ern2004-oo,
Hsiao2008-xb, Sauter2011-an}

\section{Preliminary Definitions}

@@

@@

\begin{comment}
\section{Mixed Blabbering}

>>>>

For brevity's sake, some results and derivations have been omitted from the main
body of the thesis. I collect the ones I judge most relevant in this Appendix.
Section \ref{sec:T1-cc} presents the $T_1$-transformed form of the
\acrshort{CCS}, \acrshort{CCSD} and \acrshort{CCSDT} Lagrangians, the corresponding amplitudes
and multipliers equations and the one-electron operators expectation values.

Basic results in the fields of functional analysis and boundary integral
equations that were omitted from Chapter \ref{ch:CSM} are presented here.
No proofs or examples are given, the interested reader is referred to the
monographs by \citeauthor{Ern2004-oo}, \citeauthor{Hsiao2008-xb} and
\citeauthor{Sauter2011-an}\autocite{Ern2004-oo, Hsiao2008-xb, Sauter2011-an}

\section{The \texorpdfstring{$T_1$}{T1}-Transformation}\label{sec:T1-cc}

$T_1$ is a one-electron operator, carrying out a similarity transformation of
an operator $O$ by means of its exponential mapping will preserve the particle
rank of $O$.\autocite{Koch1994-vr}
Starting from the arbitrary order Lagrangian \eqref{eq:cc-lagrangian},
we want to derive their $T_1$-transformed
expressions.
For the \acrshort{CCS} model, this is straightforward.
The Lagrangian is:
\begin{equation}\label{eq:ccs-lagrangian}
 \mathcal{L}(\tamp{}, \tbar{})_\mathrm{CCS}
 =
  E_0
  + \tbar{\mu_1}\denom{\mu_1}\tamp{\mu_1}
  + \braket{\mathrm{HF} | \check{\Phi} | \mathrm{HF}}
  + \braket{\tbar{1} | \check{\Phi} | \mathrm{HF}}
\end{equation}
the governing equations:
\begin{subequations}\label{eq:ccs-equations}
  \begin{align}
   \denom{\mu_1}\tamp{\mu_1} &+ \braket{\mu_1 | \check{\Phi} | \mathrm{HF}}
             = 0 \label{eq:ccs-amplitudes}\\
    \denom{\mu_1}\tbar{\mu_1} &+
    \braket{\mathrm{HF} | \commutator{\check{\Phi}}{\cluster{\mu_1}} | \mathrm{HF}} +
    \braket{\tbar{1} |
    \commutator{\check{\Phi}}{\cluster{\mu_1}} | \mathrm{HF}}
             = 0 \label{eq:ccs-multipliers}
  \end{align}
\end{subequations}

When including higher order excitations we will seek simplifications in the commutator expansions
by employing the following result:\autocite{Helgaker2000-tz}
\begin{lemma}[Excitation ranks manifold]\label{lemma:exc-rank}
The $k$-fold nested commutator of a \emph{particle rank} $m_O$ operator $O$ with
cluster operators $T_{n_i}$ of rank $n_i$ acting on the reference determinant:
\begin{equation}
  \commutator{\commutator{\commutator{O}{T_{n_1}}}{\ldots}}{T_{n_k}} \ket{\mathrm{HF}}
\end{equation}
generates a linear combination of determinants with excitation ranks $v$ in the
range:
\begin{equation}
  \sum_{i=1}^k n_i - m_O \leq v \leq \sum_{i=1}^k n_i + m_O -k
\end{equation}
\end{lemma}
\begin{corollary}[Excited states overlaps]
The overlap of a determinant $\mu_n$ with excitation rank $n$ onto the linear
combination of determinants generated by the $k$-fold nested commutator of an
operator $O$ with particle rank $m_O$ with cluster operators $T_{n_i}$ of rank $n_i$ is
nonzero if and only if the sum of cluster operator ranks satisfies:
\begin{equation}
  n - m_O + k \leq \sum_{i=1}^k n_i \leq n + m_O
\end{equation}
In compact form:
\begin{equation}
  \braket{\mu_n |
  \commutator{\commutator{\commutator{O}{T_{n_1}}}{\ldots}}{T_{n_k}} | \mathrm{HF}} \neq 0
  \Longleftrightarrow
  n - m_O + k \leq \sum_{i=1}^k n_i \leq n + m_O
\end{equation}
\end{corollary}

The \acrshort*{CCSD} Lagrangian is then:
\begin{equation}\label{eq:ccsd-lagrangian}
 \begin{aligned}
 \mathcal{L}(\tamp{}, \tbar{})_\mathrm{CCSD}
&=
 E_0
 + \sum_{u=1}^{2}\tbar{\mu_u}\denom{\mu_u}\tamp{\mu_u}
 + \braket{\mathrm{HF} | \check{\Phi} + \commutator{\check{\Phi}}{T_2} | \mathrm{HF}} \\
 &+ \braket{\tbar{1} | \check{\Phi} + \commutator{\check{\Phi}}{T_2} | \mathrm{HF}} \\
 &+ \braket{\tbar{2} | \check{\Phi} + \commutator{\check{\Phi}}{T_2}
   +\frac{1}{2}\commutator{\commutator{\check{\Phi}}{T_2}}{T_2} | \mathrm{HF}}
  \end{aligned}
\end{equation}
with the governing equations:
\begin{subequations}\label{eq:ccsd-equations}
  \begin{align}
   \denom{\mu_1}\tamp{\mu_1} &+ \braket{\mu_1 | \check{\Phi} + \commutator{\check{\Phi}}{T_2} | \mathrm{HF}}
             = 0 \label{eq:ccsd-singles-amplitudes}\\
   \denom{\mu_2}\tamp{\mu_2} &+
   \braket{\mu_2 | \check{\Phi} + \commutator{\check{\Phi}}{T_2} +\frac{1}{2}\commutator{\commutator{\check{\Phi}}{T_2}}{T_2} | \mathrm{HF}}
             = 0 \label{eq:ccsd-doubles-amplitudes}\\
    \denom{\mu_1}\tbar{\mu_1} &+
    \braket{\mathrm{HF} | \commutator{\check{\Phi}}{\cluster{\mu_1}} | \mathrm{HF}} +
    \braket{\tbar{1} |
      \commutator{\check{\Phi}}{\cluster{\mu_1}}
    + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{T_2}
    | \mathrm{HF}} \nonumber \\
   &+ \braket{\tbar{2} |
      \commutator{\check{\Phi}}{\cluster{\mu_1}}
    + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{T_2}
    | \mathrm{HF}}
             = 0 \label{eq:ccsd-singles-multipliers} \\
    \denom{\mu_2}\tbar{\mu_2} &+
    \braket{\mathrm{HF} | \commutator{\check{\Phi}}{\cluster{\mu_2}} | \mathrm{HF}} +
    \braket{\tbar{1} |
      \commutator{\check{\Phi}}{\cluster{\mu_2}}
    | \mathrm{HF}} \nonumber \\
   &+ \braket{\tbar{2} |
      \commutator{\check{\Phi}}{\cluster{\mu_2}}
    + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_2}}}{T_2}
    | \mathrm{HF}}
             = 0 \label{eq:ccsd-doubles-multipliers}
  \end{align}
\end{subequations}
Equations \eqref{eq:ccsd-doubles-amplitudes},
\eqref{eq:ccsd-singles-multipliers} and \eqref{eq:ccsd-doubles-multipliers} can
be compared with the corresponding \acrshort{CC2} equations
\eqref{eq:cc2-doubles-amplitudes}, \eqref{eq:cc2-singles-multipliers} and
\eqref{eq:cc2-doubles-multipliers}, respectively, to highlight which terms were neglected in
constructing the iterative method for the approximate inclusion of double excitations.

The \acrshort{CCSDT} Lagrangian is built in the same exact manner:
\begin{equation}\label{eq:ccsdt-lagrangian}
 \begin{aligned}
 \mathcal{L}(\tamp{}, \tbar{})_\mathrm{CCSDT}
&=
 E_0
 + \sum_{u=1}^{3}\tbar{\mu_u}\denom{\mu_u}\tamp{\mu_u}
 + \braket{\mathrm{HF} | \check{\Phi} + \commutator{\check{\Phi}}{T_2} | \mathrm{HF}} \\
 &+ \braket{\tbar{1} | \check{\Phi} + \commutator{\check{\Phi}}{T_2} + \commutator{\check{\Phi}}{T_3} | \mathrm{HF}} \\
 &+ \braket{\tbar{2} | \check{\Phi} + \commutator{\check{\Phi}}{T_2}
   +\frac{1}{2}\commutator{\commutator{\check{\Phi}}{T_2}}{T_2} + \commutator{\check{\Phi}}{T_3} | \mathrm{HF}} \\
 &+ \braket{\tbar{3} | \commutator{\check{\Phi}}{T_2}
   + \frac{1}{2}\commutator{\commutator{\check{\Phi}}{T_2}}{T_2} + \commutator{\check{\Phi}}{T_3}
   + \commutator{\commutator{\check{\Phi}}{T_2}}{T_3}
   | \mathrm{HF}}
  \end{aligned}
\end{equation}
Notice that the reference expectation value of the similarity transformed
fluctuation potential $\overline{\Phi}$ is unchanged with respect to the
expression in the \acrshort{CCSD} Lagrangian.
This is a consequence of the well-known fact that the \acrshort*{CC} energy can
be expressed purely in terms of $T_1$ and $T_2$:
\begin{equation}
E_\mathrm{CC} = \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
=
\braket{\mathrm{HF} | H(1 + T_1 + T_2 + T_1^2) | \mathrm{HF}}
\end{equation}
The \acrshort{CCSDT} amplitude equations are:
\begin{subequations}\label{eq:ccsdt-amplitudes}
  \begin{align}
   \denom{\mu_1}\tamp{\mu_1} &+ \braket{\mu_1 |
   \check{\Phi} + \commutator{\check{\Phi}}{T_2} + \commutator{\check{\Phi}}{T_3}
   | \mathrm{HF}}
             = 0 \label{eq:ccsdt-singles-amplitudes}\\
   \denom{\mu_2}\tamp{\mu_2} &+ \braket{\mu_2 |
   \check{\Phi} + \commutator{\check{\Phi}}{T_2}
   + \frac{1}{2}\commutator{\commutator{\check{\Phi}}{T_2}}{T_2} + \commutator{\check{\Phi}}{T_3}
   | \mathrm{HF}}
             = 0 \label{eq:ccsdt-doubles-amplitudes} \\
   \denom{\mu_3}\tamp{\mu_3} &+ \braket{\mu_3 |
   \commutator{\check{\Phi}}{T_2}
   + \frac{1}{2}\commutator{\commutator{\check{\Phi}}{T_2}}{T_2} + \commutator{\check{\Phi}}{T_3}
   + \commutator{\commutator{\check{\Phi}}{T_2}}{T_3}
   | \mathrm{HF}}
             = 0 \label{eq:ccsdt-triples-amplitudes}
  \end{align}
\end{subequations}
while the multipliers are determined by solving the following:
\begin{subequations}\label{eq:ccsdt-multipliers}
  \begin{align}
    \denom{\mu_1}\tbar{\mu_1} &+
    \braket{\mathrm{HF} | \commutator{\check{\Phi}}{\cluster{\mu_1}} | \mathrm{HF}} +
    \braket{\tbar{1} |
      \commutator{\check{\Phi}}{\cluster{\mu_1}}
    + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{T_2}
    | \mathrm{HF}} \nonumber \\
   &+ \braket{\tbar{2} |
      \commutator{\check{\Phi}}{\cluster{\mu_1}}
    + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{T_2}
    + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{T_3}
    | \mathrm{HF}} \nonumber \\
   &+ \braket{\tbar{3} |
      \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{T_2}
    + \frac{1}{2}\commutator{\commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{T_2}}{T_2}
    + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{T_3}
    | \mathrm{HF}}
             = 0 \label{eq:ccsdt-singles-multipliers} \\
    \denom{\mu_2}\tbar{\mu_2} &+
    \braket{\mathrm{HF} | \commutator{\check{\Phi}}{\cluster{\mu_2}} | \mathrm{HF}} +
    \braket{\tbar{1} |
      \commutator{\check{\Phi}}{\cluster{\mu_2}}
    | \mathrm{HF}} \nonumber \\
   &+ \braket{\tbar{2} |
      \commutator{\check{\Phi}}{\cluster{\mu_2}}
    + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_2}}}{T_2}
    | \mathrm{HF}} \nonumber \\
   &+ \braket{\tbar{3} |
      \commutator{\check{\Phi}}{\cluster{\mu_2}}
    +  \commutator{\commutator{\check{\Phi}}{\cluster{\mu_2}}}{T_2}
    +  \commutator{\commutator{\check{\Phi}}{\cluster{\mu_2}}}{T_3}
    | \mathrm{HF}}
             = 0 \label{eq:ccsdt-doubles-multipliers} \\
    \denom{\mu_3}\tbar{\mu_3} &+
    \braket{\tbar{1} |
      \commutator{\check{\Phi}}{\cluster{\mu_3}}
    | \mathrm{HF}}
    + \braket{\tbar{2} |
      \commutator{\check{\Phi}}{\cluster{\mu_3}}
    | \mathrm{HF}} \nonumber \\
   &+ \braket{\tbar{3} |
      \commutator{\check{\Phi}}{\cluster{\mu_3}}
    +  \commutator{\commutator{\check{\Phi}}{\cluster{\mu_3}}}{T_2}
    | \mathrm{HF}}
             = 0 \label{eq:ccsdt-triples-multipliers}
  \end{align}
\end{subequations}
A comparison of equations \eqref{eq:ccsdt-triples-amplitudes},
\eqref{eq:ccsdt-singles-multipliers}, \eqref{eq:ccsdt-doubles-multipliers} and
\eqref{eq:ccsdt-triples-multipliers}
with equations \eqref{eq:cc3-triples-amplitudes},
\eqref{eq:cc3-singles-multipliers}, \eqref{eq:cc3-doubles-multipliers} and
\eqref{eq:cc3-triples-multipliers}, respectively, serves to highlight which terms were
neglected in constructing the \acrshort{CC3} method.

\section{Coupled Cluster Expectation Values}\label{sec:CC-expval}

A similar analysis in terms of $T_1$-transformed operators can be given for the
expectation value of one-electron operators, Eq.
\eqref{eq:cc-expectation-value}.
First of all, by virtue of the cluster commutation
condition,\autocite{Helgaker2000-tz} the commutator expansion of the similarity
transformation of a general one-electron operator truncates after the twofold
nested commutator:
\begin{equation}
\overline{O} = O
  + \commutator{O}{T}
  + \frac{1}{2}\commutator{\commutator{O}{T}}{T}
\end{equation}
where $T$ is the complete cluster operator. However, due to Lemma
\ref{lemma:exc-rank}, some of the cluster operators will not contribute to the
expectation value.

For a general truncation level $\mathcal{M}$ one has:
\begin{equation}
  \begin{aligned}
  O(\tamp{}, \tbar{}) &=
  \braket{\mathrm{HF} | O + \BCHfirst{O}{T} + \BCHsecond{O}{T} | \mathrm{HF}} \\
  &+ \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | O + \BCHfirst{O}{T} +
  \BCHsecond{O}{T} | \mathrm{HF}},
  \end{aligned}
\end{equation}
and by virtue of Lemma \ref{lemma:exc-rank} and its Corollary:
\begin{equation}
\begin{aligned}
  O(\tamp{}, \tbar{})
  &=
  \braket{\mathrm{HF} | O + \BCHfirst{O}{T_1} | \mathrm{HF}}
  + \braket{\tbar{1} | O | \mathrm{HF}} \\
  &+ \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \BCHfirst{O}{T} + \BCHsecond{O}{T} | \mathrm{HF}}
\end{aligned}
\end{equation}
Note that the singles amplitudes assume a unique role in the \acrshort{CC}
expectation value.
In the following, we give explicit expressions for the \acrshort{CCS}, \acrshort{CCSD}
and \acrshort{CCSDT} models. These results are at the basis of our developments
in Chapter \ref{ch:solvation-correlation}.
We will refer to the $T_1$-transformed expressions as \emph{dressed}, in contrast to the
\emph{bare} expressions, where the operator $O$ appears untransformed.

For the \acrshort{CCS} model, the expectation value of a one-electron operator
is simply:
\begin{equation}\label{eq:ccs-expectation-value}
\begin{aligned}
  O(\tamp{}, \tbar{})_\mathrm{CCS} &=
  \braket{\mathrm{HF} | \check{O} | \mathrm{HF} }
  + \braket{\tbar{1} | \check{O} | \mathrm{HF} } \\
  &=
  \braket{\mathrm{HF} | O + \commutator{O}{T_1} | \mathrm{HF} }
  + \braket{\tbar{1} | O + \commutator{O}{T_1}
  + \frac{1}{2}\commutator{\commutator{O}{T_1}}{T_1} | \mathrm{HF} }
\end{aligned}
\end{equation}
Adding double excitation to the manifold yields:
\begin{equation}\label{eq:ccsd-expectation-value}
  \begin{aligned}
  O(\tamp{}, \tbar{})_\mathrm{CCSD}
  &=
    O(\tamp{}, \tbar{})_\mathrm{CCS}
  + \braket{\tbar{1} | \commutator{\check{O}}{T_2} | \mathrm{HF} }
  + \braket{\tbar{2} | \commutator{\check{O}}{T_2} | \mathrm{HF} } \\
  &=
    O(\tamp{}, \tbar{})_\mathrm{CCS}
  + \braket{\tbar{1} | \commutator{O}{T_2} | \mathrm{HF} }
  + \braket{\tbar{2} | \commutator{O}{T_2}
      + \commutator{\commutator{O}{T_1}}{T_2}
  | \mathrm{HF} }
  \end{aligned}
\end{equation}
a number of terms was dropped thanks to Lemma \ref{lemma:exc-rank} and its
Corollary.
Eventually, including triples one obtains:
\begin{equation}\label{eq:ccsdt-expectation-value}
\begin{aligned}
  O(\tamp{}, \tbar{})_\mathrm{CCSDT}
  &=
    O(\tamp{}, \tbar{})_\mathrm{CCSD}
  + \braket{\tbar{2} | \commutator{\check{O}}{T_3} | \mathrm{HF} } \\
  &+ \braket{\tbar{3} | \commutator{\check{O}}{T_3}
    + \frac{1}{2}\commutator{\commutator{\check{O}}{T_2}}{T_2}
  | \mathrm{HF} } \\
  &=
    O(\tamp{}, \tbar{})_\mathrm{CCSD}
  + \braket{\tbar{2} | \commutator{O}{T_3} | \mathrm{HF} } \\
  &+ \braket{\tbar{3} | \commutator{O}{T_3}
    + \commutator{\commutator{O}{T_1}}{T_3}
    + \frac{1}{2}\commutator{\commutator{O}{T_2}}{T_2}
  | \mathrm{HF} }
\end{aligned}
\end{equation}
where Lemma \ref{lemma:exc-rank} and its corollary were again extensively employed.

The use of normal-ordered operators is common in \acrlong*{CC} theory. Both for
the dressed and bare representations, the expectation values of such operators
can be formed by simply replacing operator with its normal-ordered counterpart,
$\check{O}_\mathrm{N}$ or $O$, respectively.
This achieves elimination of the reference expectation value from the first
term in equation \eqref{eq:ccs-expectation-value}.

\section{Selected Results in Functional Analysis}

@@

\begin{comment}

@@

\section{Mixed Blabbering}
